Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-09-21 README.Rmd
Minor edits

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

@@

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

>>>>

> Roalkvam, Isak and Solheim, Steinar (2023). _`r Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

@@

> Roalkvam, Isak and Solheim, Steinar (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### 2023-09-21 analysis/paper/paper.Rmd
Minor edits

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10,000 times for each null model. The 2.5th and 97.5th percentile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model under consideration can be rejected.

@@

While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. This is also relevant for the treatment of RSPDs in Scandinavia more broadly, as we suggested that a combination of mobility patterns, pyrotechnologies and potentially limited vegetational cover in the first period of human occupation in the region largely explain the absence of radiocarbon dates from this earliest period. The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods, possibly as mediated by pyrotechnology, remains unclear, and could prove important for properly isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@crombe2014; @hiscock2016, 219].

>>>>

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE estimation with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10,000 times for each null model. The 2.5th and 97.5th percentile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model under consideration can be rejected.

@@

While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. This is also relevant for the treatment of RSPDs in Scandinavia more broadly, as we suggested that mobility patterns have relevance for the kind of pyrotechnologies that were utilised in the first period of human occupation in the region, and that the technologies utilised largely explain the absence of radiocarbon dates from this earliest period. The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods, possibly as mediated by pyrotechnology, remains unclear, and could prove important for properly isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@crombe2014; @hiscock2016, 219].
